---
title: "Pre-Process Data from Plate Reader"
subtitle: "Pretest for hydrolytic extracellular enzyme assays: Automating the creation of CSV templates for use with the plater package"
author: "Hava Blair"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: 
    toc: true 
    number_sections: true
---
\newpage

# Before you start 

### About `plater`

The R package `plater` provides helpful functions for transforming data from plate experiments into "tidy" format for analysis.  You can read more about the plater package here: (https://docs.ropensci.org/plater/).   

A pre-requisite for using the `plater` functions is getting your plate data in the correct format.  `plater` has an excellent vignette that walks through some examples of data formatting and `plater` functions step-by-step: https://cran.r-project.org/web/packages/plater/vignettes/plater-basics.html 

### Purpose of this R Markdown 

If you have data from many plates to process, it can take some time and effort to manually create templates for each plate.  In my experience, the process of opening multiple excel files and copy / pasting the data was tedious and easy to mess up if I wasn't paying attention to my clipboard and file names. 

**This script lends a hand by automating the process of preparing microplate data stored in multiple excel files into the proper format for `plater` functions.**

The script was written to ingest an arbitrary number of excel files with multiple worksheets (tabs) in each file.  It will also work for files with only 1 worksheet.  

### What you need 

- Data exported from the plate reader software (this script assumes .xlsx). The name of each tab should be a plate number or identifier, and each one should be unique. 
- Your plate layouts (map of sample names, blanks, checks on each plate)
- Your lab notebook or notes with information about any known pipetting errors or plate issues that should be flagged for quality control ("Bad Wells")
- A generic plate template.  Templates are available in the `empty_templates` sub-directory for the hydrolytic and oxidative extracellular enzyme assays regularly performed in the Gutknecht lab.  If you need to create your own template, see Step 1 in this tutorial for `plater`'s formatting requirements:  https://cran.r-project.org/web/packages/plater/vignettes/plater-basics.html
\newpage

# Setup

If you do not already have the packages below installed in R, you may need to install them first.  

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff=60))

library(tidyverse)
library(readxl)
library(glue)
library(plater)

```

# Read in excel files 

The `raw_excel_pretest` subdirectory has some test data that you can use to play around with this script before you feed in your own data.  **Don't forget to delete the test files (or move them to a different folder) before you use this script for your own data.**

You can have as many sheets (tabs) as you want in each excel file that you intend to process.  Place all the excel files you want to process in the `raw_excel_pretest` sub-directory.   Do not place additional files in this sub-directory. 

```{r read}
#fill in directory (folder) where your raw excel files are
xl_files <- dir("./raw_excel_pretest/")

# save full file path to each of your excel files
xl_paths <- paste0("./raw_excel_pretest/", xl_files)

# function to create dataframe for file path and sheet name info. Each row in this dataframe represents a single "sheet" (tab) from one of your raw excel files. You will get one dataframe per excel file in your original list.
# NOTE: 'sheet' col specified as "character" so bind_rows() below doesn't complain about unequal factor levels. 
make_df <- function(file_path){
df <- data.frame(sheet = excel_sheets(path = file_path))
df$path = file_path
df$sheet <- as.character(df$sheet)
df
}

# apply make_df to every excel file you want to process
# creates list of dfs
sheets_list <- map(xl_paths, make_df)

sheets_list

``` 


# Create single dataframe of file paths and sheet names 

We will build a simple dataframe.  Each row will contain the file path to one of our excel files and a corresponding sheet name for each sheet we want to turn into a plater template.  

```{r file-sheet-list, message=FALSE}

# bind list of dfs together into a single df:
sheets_path_df <- bind_rows(sheets_list) 
# path should be first col, sheets second col
sheets_path_df <- sheets_path_df %>% select(path, sheet)

# peek at first few rows
head(sheets_path_df)
```


# Extract the fluorescence data from your excel files

The plate reader includes extra metadata and formatting junk in the exported worksheets.  The function below extracts just the rectangle containing the data including column and row names. 

```{r extract-data, message=FALSE}
# function to read only fluorescence data from each sheet in every excel file

read_sheets <- function(path, sheet){
  
  raw_data <- read_excel(path = path, sheet = sheet, range = "B26:N34")
  raw_data %>%  
    rename(Template = ...1) %>% 
    mutate(`1` = as.character(`1`),
           `2` = as.character(`2`),
           `3` = as.character(`3`),
           `4` = as.character(`4`),
           `5` = as.character(`5`),
           `6` = as.character(`6`),
           `7` = as.character(`7`),
           `8` = as.character(`8`),
           `9` = as.character(`9`), 
           `10` = as.character(`10`), 
           `11` = as.character(`11`), 
           `12` = as.character(`12`))
}

# save the fluor data as a list column 
fluor_import <- sheets_path_df %>% 
  mutate(fluor_data = pmap(sheets_path_df, read_sheets))

```

# Create plater templates

This function reads in a template from the empty_templates sub-directory as a dataframe.  It then appends the raw data we extracted above to the empty "Data" section at the end of the template.   

```{r templates}

# function to add fluorescence data to plater template

template_fun <- function(sheet, fluor_data){
  
  #load blank template for BG/Cello (bc) plates
  blank_template_bc <- read.csv("./empty_templates/blank_template_pretest_bc.csv",
                            stringsAsFactors = FALSE,
                            blank.lines.skip = FALSE)
    
  # set col names BG/Cello (bc) plate template
  colnames(blank_template_bc) <- c("Template", "1", "2", "3", "4", "5",
                             "6", "7", "8", "9", "10", "11", "12")
  
  # load blank template for NAG/P (np) plates 
  blank_template_np <- read.csv("./empty_templates/blank_template_pretest_np.csv", stringsAsFactors = FALSE, blank.lines.skip = FALSE)
  
  # set col names for NAG/P (np) plate template
   colnames(blank_template_np) <- c("Template", "1", "2", "3", "4", "5",
                             "6", "7", "8", "9", "10", "11", "12")
   #bind template and plate data, detecting if A Plate or B Plate template needed.
  if (str_detect(sheet, "bc")){
    template <- rbind(blank_template_bc, fluor_data)
  }else{
    template <- rbind(blank_template_np, fluor_data)
  }
   
  #return completed template
  return(template)
}

```

# Create a folder (directory)

Before we use the function above to fill out our `plater` templates, we want to create a place to save them.  The code below creates a folder (directory) within the current working directory (folder where you have saved this R project). 

Note: If you are running this script multiple times (testing/troubleshooting) you may see a warning that "auto-templates" directory already exists.  It's not a problem, but note that every time you run this part of the script, the sub-directory and its contents will be over-written. 
```{r}

# create a folder (directory) called "auto-templates"

dir.create("./auto_templates_pretest/")

# plater_templates col contains completed template dfs
# export_paths col contains full path and file name for saving each completed template
incl_templates <- fluor_import %>% 
  mutate(plater_templates = map2(sheet, fluor_data, template_fun), 
         export_paths = as.character(glue("./auto_templates_pretest/plater-{sheet}.csv")))

```


# Save completed templates as CSVs

Now that we have our completed templates in dataframe form, we need to write them to CSV files and save them for future analysis.  

```{r save-data, message=FALSE}

# function to save each template as a CSV using the specified export path
save_templates_fun <- function(plater_templates, export_paths){
  write_csv(plater_templates, export_paths, col_names = TRUE)
}

# write all templates to CSV, save in auto-templates sub-directory
save_templates <- incl_templates %>% 
  mutate(saved = walk2(plater_templates, export_paths, save_templates_fun))

```

# Initial quality control

Now that you have one CSV file template saved for each plate, you should consult your lab notes and flag bad wells in the "Bad_wells" section of the appropriate template with the word 'bad'.  Additional quality control will be performed in the analysis script.  

# Additional Improvements 


```{r read-test}

test_dir <- dir("./auto_templates_pretest/")

test_paths <- paste0("./auto_templates_pretest/", test_dir)

read_plates(test_paths)

```


